---
title: "Scenario Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scenario Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Starting From The Kirotshe Fit

Scenario analysis should start from a model that can reproduce the outbreak we
are using as the baseline. The fitting vignette saves the stochastic pMCMC fit
for Kirotshe as package example data, so here we read that object rather than
refitting the model.

```{r}
library(chlaa)
library(ggplot2)
library(dplyr)

fit_obj <- readRDS(system.file(
  "extdata", "kirotshe_particle_fit.rds",
  package = "chlaa",
  mustWork = TRUE
))

fit <- fit_obj$fit
base_pars <- fit_obj$pars
observed <- fit_obj$observed
burnin <- fit_obj$burnin

pars_fit <- chlaa_update_from_fit(
  fit = fit,
  pars = base_pars,
  draw = "median",
  burnin = burnin
)
```

The posterior predictive check shows the fitted distribution for weekly
reported cases. The points are the observed Kirotshe data; the ribbons and line
come from posterior draws.

```{r}
fit_fc <- chlaa_forecast_from_fit(
  fit = fit,
  pars = base_pars,
  time = observed$time,
  vars = "inc_symptoms_weekly",
  include_cases = TRUE,
  obs_interval = 7,
  obs_model = "nbinom",
  n_draws = 80,
  burnin = burnin,
  seed = 11,
  dt = 1
)

chlaa_plot_forecast(
  fit_fc,
  var = "cases",
  data = observed,
  data_y = "cases"
)
```

## Counterfactual Scenarios

The baseline scenario is the fitted outbreak with the recorded intervention
timings. We compare it with one no-intervention counterfactual and two
anticipatory-action examples. The AA examples move WASH and care earlier, and
one adds a single-dose vaccination campaign.

```{r}
horizon <- max(observed$time) + 182
scenario_time <- seq(7, horizon, by = 7)
trigger_time <- min(observed$time[observed$cases >= 50])
response_end <- horizon + 1
campaign_days <- 28
vaccine_doses <- floor(0.20 * pars_fit$N)

no_intervention <- list(
  chlor_start = 0, chlor_end = 0, chlor_effect = 0,
  hyg_start = 0, hyg_end = 0, hyg_effect = 0,
  lat_start = 0, lat_end = 0, lat_effect = 0,
  cati_start = 0, cati_end = 0, cati_effect = 0,
  orc_start = 0, orc_end = 0, orc_capacity = 0,
  ctc_start = 0, ctc_end = 0, ctc_capacity = 0,
  vax1_start = 0, vax1_end = 0, vax1_doses_per_day = 0, vax1_total_doses = 0,
  vax2_start = 0, vax2_end = 0, vax2_doses_per_day = 0, vax2_total_doses = 0
)

aa_response <- list(
  chlor_start = trigger_time, chlor_end = response_end, chlor_effect = pars_fit$chlor_effect,
  hyg_start = trigger_time, hyg_end = response_end, hyg_effect = pars_fit$hyg_effect,
  lat_start = trigger_time, lat_end = response_end, lat_effect = max(pars_fit$lat_effect, 0.10),
  cati_start = trigger_time, cati_end = response_end, cati_effect = pars_fit$cati_effect,
  orc_start = trigger_time, orc_end = response_end, orc_capacity = pars_fit$orc_capacity,
  ctc_start = trigger_time, ctc_end = response_end, ctc_capacity = pars_fit$ctc_capacity,
  vax1_start = 0, vax1_end = 0, vax1_doses_per_day = 0, vax1_total_doses = 0,
  vax2_start = 0, vax2_end = 0, vax2_doses_per_day = 0, vax2_total_doses = 0
)

aa_vaccination <- modifyList(aa_response, list(
  vax1_start = trigger_time + 14,
  vax1_end = trigger_time + 14 + campaign_days,
  vax1_total_doses = vaccine_doses,
  vax1_doses_per_day = vaccine_doses / campaign_days
))

scenarios <- list(
  chlaa_scenario("no_interventions", no_intervention),
  chlaa_scenario("aa_response", aa_response),
  chlaa_scenario("aa_response_plus_vaccine", aa_vaccination)
)

vapply(scenarios, `[[`, character(1), "name")
```

## Posterior Scenario Forecasts

`chlaa_forecast_scenarios_from_fit()` uses the same posterior draws for each
scenario, then reports absolute forecasts and paired differences against the
baseline. That pairing matters because it removes some Monte Carlo noise from
scenario differences.

```{r}
scenario_fc <- chlaa_forecast_scenarios_from_fit(
  fit = fit,
  pars = base_pars,
  scenarios = scenarios,
  baseline_name = "fitted_response",
  time = scenario_time,
  vars = c("inc_symptoms_weekly", "cum_symptoms", "cum_deaths"),
  include_cases = TRUE,
  obs_interval = 7,
  n_draws = 60,
  burnin = burnin,
  seed = 12,
  dt = 1
)
```

```{r}
chlaa_plot_scenario_forecasts(
  scenario_fc,
  var = "cases",
  type = "absolute",
  data = observed,
  data_y = "cases"
)
```

```{r}
chlaa_plot_scenario_forecasts(
  scenario_fc,
  var = "cum_deaths",
  type = "difference",
  include_baseline = FALSE
)
```

## Decision Summary

For compact decision tables we simulate the same scenarios using the posterior
median parameter set. This is faster and easier to inspect, while the forecast
plots above show posterior uncertainty.

```{r}
scenario_runs <- chlaa_run_scenarios(
  pars = pars_fit,
  scenarios = c(list(chlaa_scenario("fitted_response", list())), scenarios),
  time = scenario_time,
  n_particles = 50,
  dt = 1,
  seed = 13
)

chlaa_scenario_summary(
  scenario_runs,
  baseline = "fitted_response",
  incidence_var = "inc_symptoms_weekly"
)
```

```{r}
chlaa_compare_scenarios(
  scenario_runs,
  baseline = "fitted_response",
  include_econ = TRUE,
  wtp = 1500
)
```

The no-intervention scenario is useful for estimating how much the recorded
response may have changed the outbreak. The AA scenarios ask a different
question: what changes if similar response levers are triggered earlier, with
or without vaccination.
