---
title: "Fitting"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.path = "fitting/",
  message = FALSE,
  warning = FALSE
)

library(chlaa)
library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(tibble)
library(tidyr)

if (!requireNamespace("coda", quietly = TRUE)) {
  stop("The coda package is needed for ESS and R-hat diagnostics.")
}

theme_set(theme_minimal(base_size = 12))
```

# Overview

This vignette fits the cholera transmission model to weekly case counts from
Kirotshe health zone. The data are weekly IDSR counts, so the observation model
is a count model rather than a model for per-capita rates.

The aim is to build up a stable pMCMC workflow for a deliberately small model.
We fix the intervention dates and most natural-history parameters, set
person-to-person transmission to zero, and estimate four quantities:

* `trans_prob`: environmental transmission probability;
* `reporting_rate`: the fraction of symptomatic infections reported as cases;
* `obs_size`: the Negative Binomial overdispersion parameter;
* `E0`: the initial exposed population.

The workflow has two important features. First, we use deterministic-filter
chains to learn the geometry of the posterior before running the particle
filter. Second, we run several chains from dispersed starting points, because
agreement between chains is one of the simplest checks that the fit is not just
following its starting value.

# Weekly observations

The model has both daily and weekly incidence accumulators. Daily workflows use
`inc_symptoms`; weekly workflows use `inc_symptoms_weekly`. For weekly data the
likelihood is

\[
\text{cases}_t \sim \text{Negative Binomial}\{\mu_t, \text{obs_size}\},
\qquad
\mu_t = \text{reporting_rate} \times \text{inc_symptoms_weekly}_t.
\]

The first weekly observation is placed at `time = 7`, so it is compared with
incidence accumulated over days 0 to 7. This is why `chlaa_fit_pmcmc()` starts
weekly filters at day 0 when observations are at `7, 14, ...`.

# Kirotshe data

The package includes a small Kirotshe-only extract for this workflow. It
contains the weekly outbreak counts and the intervention dates/effects needed
for the fit; the larger analysis datasets remain outside the package build.
Population size comes from the weekly IDSR extract and is passed to the process
model as `N`.

```{r load-data}
repo_file <- function(...) {
  candidates <- c(file.path(...), file.path("..", ...))
  out <- candidates[file.exists(candidates)][1]
  if (is.na(out)) stop("Could not find file: ", file.path(...), call. = FALSE)
  out
}

repo_output_file <- function(...) {
  roots <- c(".", "..")
  root <- roots[file.exists(file.path(roots, "DESCRIPTION"))][1]
  if (is.na(root)) stop("Could not find repository root", call. = FALSE)
  file.path(root, ...)
}

extdata_file <- function(...) {
  out <- system.file("extdata", ..., package = "chlaa")
  if (nzchar(out)) return(out)
  repo_file("inst", "extdata", ...)
}

hz_name <- "kirotshe"

kirotshe <- read_csv(extdata_file("kirotshe_weekly_cases.csv"), show_col_types = FALSE) |>
  mutate(date = as.Date(date))

kirotshe_meta <- read_csv(
  extdata_file("kirotshe_interventions.csv"),
  col_types = cols(.default = col_character()),
  na = c("NA", "")
)

outbreak_start <- as.Date(kirotshe_meta$outbreak_start)
outbreak_end <- as.Date(kirotshe_meta$outbreak_end)

kirotshe |>
  select(date, time, cases, population, cases_pop) |>
  head()
```

The model time is a consecutive weekly index. If reporting weeks are missing in
a formal analysis, decide before fitting whether those weeks should be padded or
whether the observed time gaps should be preserved.

```{r observed-data-plot}
ggplot(kirotshe, aes(date, cases)) +
  geom_col(width = 6, fill = "grey65") +
  labs(
    x = NULL,
    y = "Weekly reported cases",
    title = "Kirotshe IDSR outbreak window"
  )
```

# Interventions and fixed parameters

The parameter file stores intervention dates as calendar dates. The model uses
day offsets relative to the outbreak start.

```{r intervention-days}
date_to_day <- function(x, origin) {
  d <- as.Date(x)
  if (length(d) == 0 || is.na(d)) 0L else as.integer(d - origin)
}

num_or <- function(x, default = 0) {
  y <- suppressWarnings(as.numeric(x))
  if (length(y) == 0 || is.na(y)) default else y
}

intervention_days <- list(
  orc_start = date_to_day(kirotshe_meta$orc_start, outbreak_start),
  orc_end = date_to_day(kirotshe_meta$orc_end, outbreak_start),
  ctc_start = date_to_day(kirotshe_meta$ctc_start, outbreak_start),
  ctc_end = date_to_day(kirotshe_meta$ctc_end, outbreak_start),
  chlor_start = date_to_day(kirotshe_meta$chlor_start, outbreak_start),
  chlor_end = date_to_day(kirotshe_meta$chlor_end, outbreak_start),
  chlor_effect = num_or(kirotshe_meta$chlor_effect),
  hyg_start = date_to_day(kirotshe_meta$hyg_start, outbreak_start),
  hyg_end = date_to_day(kirotshe_meta$hyg_end, outbreak_start),
  hyg_effect = num_or(kirotshe_meta$hyg_effect),
  cati_start = date_to_day(kirotshe_meta$cati_start, outbreak_start),
  cati_end = date_to_day(kirotshe_meta$cati_end, outbreak_start),
  cati_effect = num_or(kirotshe_meta$cati_effect),
  lat_start = date_to_day(kirotshe_meta$lat_start, outbreak_start),
  lat_end = date_to_day(kirotshe_meta$lat_end, outbreak_start),
  lat_effect = num_or(kirotshe_meta$lat_effect)
)

tibble(
  quantity = names(intervention_days),
  value = unlist(intervention_days)
)
```

For this example we use one small parameter factory. Everything not listed in
the function arguments is fixed across the fits.

```{r baseline-parameters}
make_kirotshe_pars <- function(trans_prob,
                               reporting_rate,
                               obs_size,
                               E0) {
  do.call(
    chlaa_parameters,
    c(
      list(
        N = kirotshe$population[1],
        contact_rate = 0,
        trans_prob = trans_prob,
        E0 = E0,
        Sev0 = 0,
        M0 = 0,
        C0 = 0,
        reporting_rate = reporting_rate,
        obs_size = obs_size,
        seek_severe = 0.4,
        fatality_treated = 0.001,
        fatality_untreated = 0.0043
      ),
      intervention_days
    )
  )
}
```

# Simulated weekly data

We first make a synthetic dataset on the Kirotshe time grid. This gives us a
known truth for checking whether the workflow can recover the parameters before
we fit the real outbreak.

```{r synthetic-data}
truth_pars <- make_kirotshe_pars(
  trans_prob = 8.5e-4,
  reporting_rate = 0.35,
  obs_size = 120,
  E0 = 80
)

truth <- chlaa_simulate(
  pars = truth_pars,
  time = kirotshe$time,
  n_particles = 1,
  seed = 1,
  dt = 1,
  deterministic = TRUE
)

set.seed(1)
synthetic_weekly <- kirotshe |>
  transmute(
    time,
    date,
    population,
    inc_symptoms_truth = truth$inc_symptoms_weekly,
    mu_cases = truth_pars$reporting_rate * inc_symptoms_truth,
    cases = rnbinom(n(), mu = pmax(mu_cases, 0.01), size = truth_pars$obs_size)
  )

natural_fit_names <- c("trans_prob", "reporting_rate", "obs_size", "E0")
truth_vec <- unlist(truth_pars[natural_fit_names])
truth_values <- tibble(parameter = names(truth_vec), truth = as.numeric(truth_vec))

truth_values
```

```{r synthetic-data-plot}
ggplot() +
  geom_col(data = synthetic_weekly, aes(date, cases), width = 6, fill = "grey70") +
  geom_line(data = synthetic_weekly, aes(date, mu_cases), colour = "#238b45", linewidth = 0.8) +
  labs(
    x = NULL,
    y = "Weekly reported cases",
    title = "Synthetic weekly data",
    subtitle = "Bars are noisy observations; the green line is the observation mean"
  )
```

# Diagnostic helpers

The next few helpers are used to summarise the chains. The code is intentionally
short: fitting itself is handled by `chlaa_fit_pmcmc()`, including multiple
chains through `n_chains` and `chain_pars`.

```{r run-settings}
n_chains <- 3L
pilot_steps <- 5000L
deterministic_steps <- 20000L
particle_steps <- 5000L
particle_count <- 50L
fit_data_synthetic <- synthetic_weekly |> select(time, cases)
fit_data_real <- kirotshe |> select(time, cases)
```

Effective sample size, or ESS, translates an autocorrelated MCMC chain into the
approximate number of independent samples it contains. A chain can save 3,750
post-burn-in draws and still have an ESS of 50 if it moves slowly. ESS is
therefore a measure of information, not just storage.

```{r diagnostics-tables}
draws_wide <- function(fit, burnin = 0.25, scale = c("sampled", "natural")) {
  scale <- match.arg(scale)
  chlaa_fit_trace(fit, burnin = burnin, scale = scale) |>
    pivot_wider(names_from = parameter, values_from = value) |>
    arrange(chain, iteration)
}

acceptance_summary <- function(fit, stage, burnin = 0.25) {
  chlaa_fit_report(fit, burnin = burnin)$acceptance_by_chain |>
    mutate(stage = stage, .before = 1)
}

ess_summary <- function(fit, stage, burnin = 0.25, parameters = NULL) {
  dr <- draws_wide(fit, burnin = burnin, scale = "sampled")
  if (is.null(parameters)) {
    parameters <- setdiff(names(dr), c("chain", "iteration"))
  }

  dr |>
    group_split(chain) |>
    map_dfr(function(d) {
      ess <- coda::effectiveSize(as.matrix(d[, parameters, drop = FALSE]))
      tibble(chain = d$chain[[1]], parameter = names(ess), ess = as.numeric(ess))
    }) |>
    mutate(stage = stage, .before = 1)
}

rhat_summary <- function(fit, stage, burnin = 0.25, parameters = NULL) {
  dr <- draws_wide(fit, burnin = burnin, scale = "sampled")
  if (is.null(parameters)) {
    parameters <- setdiff(names(dr), c("chain", "iteration"))
  }

  chains <- dr |>
    group_split(chain) |>
    map(\(d) coda::mcmc(as.matrix(d[, parameters, drop = FALSE])))

  psrf <- coda::gelman.diag(
    coda::mcmc.list(chains),
    autoburnin = FALSE,
    multivariate = FALSE
  )$psrf

  tibble(
    stage = stage,
    parameter = rownames(psrf),
    rhat = psrf[, "Point est."],
    rhat_upper = psrf[, "Upper C.I."]
  )
}
```

Posterior summaries are easier to read on the natural scale. For the synthetic
example we can also record whether the true value lies in the central 95%
posterior interval.

```{r summary-helper}
parameter_summary <- function(fit, stage, burnin = 0.25, truth = NULL) {
  out <- draws_wide(fit, burnin = burnin, scale = "natural") |>
    select(all_of(natural_fit_names)) |>
    pivot_longer(everything(), names_to = "parameter", values_to = "value") |>
    group_by(parameter) |>
    summarise(
      q025 = quantile(value, 0.025),
      median = median(value),
      q975 = quantile(value, 0.975),
      .groups = "drop"
    ) |>
    mutate(stage = stage, .before = 1)

  if (!is.null(truth)) {
    out <- out |>
      left_join(tibble(parameter = names(truth), truth = as.numeric(truth)), by = "parameter") |>
      mutate(covers_truth = q025 <= truth & truth <= q975)
  }

  out
}
```

The proposal covariance learned from a pilot run must be positive definite. The
`make_pd()` helper symmetrises the covariance matrix, floors very small or
negative eigenvalues, and reconstructs a valid covariance matrix. This prevents
numerical failures when the empirical covariance is nearly singular.

```{r proposal-helper}
make_pd <- function(x, min_eig = 1e-10) {
  x <- (x + t(x)) / 2
  eig <- eigen(x, symmetric = TRUE)
  eig$values <- pmax(eig$values, min_eig)
  out <- eig$vectors %*% diag(eig$values, nrow = length(eig$values)) %*% t(eig$vectors)
  dimnames(out) <- dimnames(x)
  out
}

proposal_from_fit <- function(fit, burnin = 0.25, scale = 1) {
  dr <- draws_wide(fit, burnin = burnin, scale = "sampled")
  theta <- as.matrix(dr[, setdiff(names(dr), c("chain", "iteration")), drop = FALSE])
  make_pd(cov(theta) * (2.38^2 / ncol(theta)) * scale)
}
```

For particle pMCMC we start each chain from the median of the corresponding
deterministic chain.

```{r start-helper}
pars_from_theta <- function(theta, template_pars, packer) {
  theta <- as.numeric(theta)
  names(theta) <- packer[["names"]]()
  unpacked <- packer[["unpack"]](theta)
  fixed_names <- names(packer[["inputs"]]()$fixed)
  update_names <- setdiff(names(unpacked), fixed_names)

  out <- template_pars
  for (nm in intersect(update_names, names(out))) {
    out[[nm]] <- unpacked[[nm]]
  }
  out
}

chain_median_starts <- function(fit, template_pars, burnin = 0.25) {
  dr <- draws_wide(fit, burnin = burnin, scale = "sampled")
  packer <- attr(fit, "packer", exact = TRUE)
  fit_names <- packer[["names"]]()

  dr |>
    group_split(chain) |>
    map(function(d) {
      theta <- vapply(fit_names, \(nm) median(d[[nm]]), numeric(1))
      pars_from_theta(theta, template_pars, packer)
    })
}
```

The posterior predictive plots use posterior draws from all chains after
burn-in.

```{r ppc-helper}
plot_case_fit <- function(fit, observed, title, seed, burnin = 0.25) {
  fc <- chlaa_forecast_from_fit(
    fit = fit,
    time = observed$time,
    vars = "inc_symptoms_weekly",
    include_cases = TRUE,
    obs_model = "nbinom",
    n_draws = 200,
    burnin = burnin,
    seed = seed,
    dt = 1,
    deterministic = FALSE
  )

  fit_cases <- fc |>
    filter(variable == "cases") |>
    left_join(observed |> select(time, date), by = "time")

  ggplot() +
    geom_ribbon(
      data = fit_cases,
      aes(date, ymin = q0p025, ymax = q0p975),
      fill = "#6baed6",
      alpha = 0.25
    ) +
    geom_ribbon(
      data = fit_cases,
      aes(date, ymin = q0p25, ymax = q0p75),
      fill = "#6baed6",
      alpha = 0.45
    ) +
    geom_line(data = fit_cases, aes(date, q0p5), colour = "#08519c", linewidth = 0.8) +
    geom_point(data = observed, aes(date, cases), size = 1.6) +
    labs(x = NULL, y = "Weekly reported cases", title = title)
}
```

# Natural-scale random walk

It is tempting to fit directly on the scale of the parameters. The starting
values below are deliberately dispersed. The synthetic truth is included so the
movement needed from each chain is visible.

```{r raw-starts}
make_raw_start <- function(trans_prob, reporting_rate, obs_size, E0) {
  make_kirotshe_pars(
    trans_prob = trans_prob,
    reporting_rate = reporting_rate,
    obs_size = obs_size,
    E0 = E0
  )
}

raw_starts <- list(
  make_raw_start(trans_prob = 1.6e-3, reporting_rate = 0.12, obs_size = 20, E0 = 35),
  make_raw_start(trans_prob = 4.0e-4, reporting_rate = 0.70, obs_size = 200, E0 = 200),
  make_raw_start(trans_prob = 1.2e-3, reporting_rate = 0.25, obs_size = 45, E0 = 120)
)

bind_rows(
  tibble(row = "truth", parameter = names(truth_vec), value = as.numeric(truth_vec)),
  imap_dfr(raw_starts, \(pars, chain) tibble(
    row = paste0("chain_", chain),
    parameter = natural_fit_names,
    value = unlist(pars[natural_fit_names])
  ))
) |>
  pivot_wider(names_from = parameter, values_from = value)
```

The raw-scale prior and packer estimate the same four model parameters directly.
This short deterministic run is not used for the final fit. It is a diagnostic
contrast.

```{r raw-setup}
raw_prior <- monty::monty_dsl({
  trans_prob ~ Uniform(1e-4, 1e-2)
  reporting_rate ~ Uniform(0.05, 0.8)
  obs_size ~ Uniform(1, 300)
  E0 ~ Uniform(5, 2000)
}, gradient = FALSE)

raw_packer <- function(pars) {
  fixed <- pars[setdiff(names(pars), natural_fit_names)]
  monty::monty_packer(scalar = natural_fit_names, fixed = fixed)
}

raw_proposal <- c(
  trans_prob = 1e-8,
  reporting_rate = 2e-4,
  obs_size = 9,
  E0 = 400
)
```

```{r raw-fit, message=FALSE}
raw_fit <- chlaa_fit_pmcmc(
  data = fit_data_synthetic,
  pars = raw_starts[[1]],
  chain_pars = raw_starts,
  n_chains = n_chains,
  n_particles = 1,
  n_steps = pilot_steps,
  seed = 91,
  prior = raw_prior,
  packer = raw_packer(raw_starts[[1]]),
  proposal_var = raw_proposal,
  obs_interval = 7,
  time_start = 0,
  deterministic = TRUE
)
```

The raw-scale chains move, but they do not explore the posterior geometry very
efficiently. Parameters with small positive values, probabilities bounded by
zero and one, and large count-like parameters all need proposal variances on
very different scales.

```{r raw-diagnostics}
acceptance_summary(raw_fit, "raw deterministic", burnin = 0.25)
```

```{r raw-ess-rhat}
ess_summary(raw_fit, "raw deterministic", burnin = 0.25)
rhat_summary(raw_fit, "raw deterministic", burnin = 0.25)
```

```{r raw-trace}
chlaa_plot_trace(
  raw_fit,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural"
)
```

```{r raw-pairs, fig.height=7}
chlaa_plot_parameter_pairs(
  raw_fit,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural",
  truth = truth_vec,
  max_points = 2500
)
```

# Transformed parameterisation

The final workflow samples transformed coordinates:

* `log_trans_prob` for a positive transmission probability;
* `logit_reporting_rate` for a probability bounded by zero and one;
* `log_obs_size` for a positive overdispersion parameter;
* `log_E0` for a positive initial condition.

The sampler moves on this transformed scale, but the plots below use the
natural scale by default.

```{r transformed-setup}
fit_names <- c(
  "log_trans_prob",
  "logit_reporting_rate",
  "log_obs_size",
  "log_E0"
)

fit_prior <- monty::monty_dsl({
  log_trans_prob ~ Uniform(-9.210340, -4.605170)       # log(c(1e-4, 1e-2))
  logit_reporting_rate ~ Uniform(-2.944439, 1.386294) # qlogis(c(0.05, 0.8))
  log_obs_size ~ Uniform(0, 5.703782)                 # log(c(1, 300))
  log_E0 ~ Uniform(1.609438, 7.600902)                # log(c(5, 2000))
}, gradient = FALSE)

add_transformed_values <- function(pars) {
  pars$log_trans_prob <- log(pars$trans_prob)
  pars$logit_reporting_rate <- qlogis(pars$reporting_rate)
  pars$log_obs_size <- log(pars$obs_size)
  pars$log_E0 <- log(pars$E0)
  pars
}

make_packer <- function(pars) {
  fixed <- pars[setdiff(names(pars), c(fit_names, natural_fit_names))]

  monty::monty_packer(
    scalar = fit_names,
    fixed = fixed,
    process = function(p) {
      list(
        trans_prob = exp(p$log_trans_prob),
        reporting_rate = plogis(p$logit_reporting_rate),
        obs_size = exp(p$log_obs_size),
        E0 = exp(p$log_E0)
      )
    }
  )
}
```

```{r transformed-starts}
make_start <- function(trans_prob, reporting_rate, obs_size, E0) {
  make_kirotshe_pars(
    trans_prob = trans_prob,
    reporting_rate = reporting_rate,
    obs_size = obs_size,
    E0 = E0
  ) |>
    add_transformed_values()
}

synthetic_starts <- list(
  make_start(trans_prob = 1.6e-3, reporting_rate = 0.12, obs_size = 20, E0 = 35),
  make_start(trans_prob = 4.0e-4, reporting_rate = 0.70, obs_size = 200, E0 = 200),
  make_start(trans_prob = 1.2e-3, reporting_rate = 0.25, obs_size = 45, E0 = 120)
)
```

# Synthetic fit

The transformed workflow begins with a deterministic pilot. The proposal is
diagonal and deliberately simple; it only needs to move enough for us to learn
the covariance structure.

```{r synthetic-pilot, message=FALSE}
pilot_proposal <- c(
  log_trans_prob = 0.02,
  logit_reporting_rate = 0.05,
  log_obs_size = 0.08,
  log_E0 = 0.08
)

synthetic_pilot <- chlaa_fit_pmcmc(
  data = fit_data_synthetic,
  pars = synthetic_starts[[1]],
  chain_pars = synthetic_starts,
  n_chains = n_chains,
  n_particles = 1,
  n_steps = pilot_steps,
  seed = 101,
  prior = fit_prior,
  packer = make_packer(synthetic_starts[[1]]),
  proposal_var = pilot_proposal,
  obs_interval = 7,
  time_start = 0,
  deterministic = TRUE
)
```

The pilot acceptance rates should not be interpreted as final diagnostics. They
tell us whether the simple proposal was able to move enough to estimate a useful
covariance.

```{r synthetic-pilot-diagnostics}
acceptance_summary(synthetic_pilot, "synthetic pilot")
```

The learned covariance is full rather than diagonal. Off-diagonal correlations
are useful here because `trans_prob`, `reporting_rate`, and `E0` can compensate
for one another.

```{r synthetic-det-proposal}
synthetic_det_proposal <- proposal_from_fit(
  synthetic_pilot,
  burnin = 0.25,
  scale = 1
)

round(cov2cor(synthetic_det_proposal), 2)
```

With the learned covariance, the deterministic chains can be run longer. The
main visual check is that chains from different starts overlap after burn-in.

```{r synthetic-deterministic-fit, message=FALSE}
synthetic_det <- chlaa_fit_pmcmc(
  data = fit_data_synthetic,
  pars = synthetic_starts[[1]],
  chain_pars = synthetic_starts,
  n_chains = n_chains,
  n_particles = 1,
  n_steps = deterministic_steps,
  seed = 201,
  prior = fit_prior,
  packer = make_packer(synthetic_starts[[1]]),
  proposal_var = synthetic_det_proposal,
  obs_interval = 7,
  time_start = 0,
  deterministic = TRUE
)
```

```{r synthetic-deterministic-tables}
acceptance_summary(synthetic_det, "synthetic deterministic")
ess_summary(synthetic_det, "synthetic deterministic")
rhat_summary(synthetic_det, "synthetic deterministic")
```

```{r synthetic-deterministic-summary}
parameter_summary(
  synthetic_det,
  "synthetic deterministic",
  truth = truth_vec
)
```

```{r synthetic-deterministic-trace}
chlaa_plot_trace(
  synthetic_det,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural"
)
```

The deterministic fit is also useful for diagnosing posterior geometry. The
dashed lines mark the synthetic truth.

```{r synthetic-det-distributions}
chlaa_plot_parameter_distributions(
  synthetic_det,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural",
  truth = truth_vec
)
```

```{r synthetic-det-pairs, fig.height=7}
chlaa_plot_parameter_pairs(
  synthetic_det,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural",
  truth = truth_vec,
  max_points = 3000
)
```

# Synthetic particle pMCMC

The particle-filter run uses the deterministic posterior medians as starting
points. We keep the covariance direction learned by the deterministic fit and
shrink the scale because the particle likelihood is noisy.

```{r synthetic-particle-starts}
synthetic_particle_starts <- chain_median_starts(
  synthetic_det,
  template_pars = synthetic_starts[[1]],
  burnin = 0.25
)
```

```{r synthetic-particle-proposal}
synthetic_particle_proposal <- proposal_from_fit(
  synthetic_det,
  burnin = 0.25,
  scale = 0.8
)
```

```{r synthetic-particle-fit, message=FALSE}
synthetic_particle <- chlaa_fit_pmcmc(
  data = fit_data_synthetic,
  pars = synthetic_particle_starts[[1]],
  chain_pars = synthetic_particle_starts,
  n_chains = n_chains,
  n_particles = particle_count,
  n_steps = particle_steps,
  seed = 301,
  prior = fit_prior,
  packer = make_packer(synthetic_particle_starts[[1]]),
  proposal_var = synthetic_particle_proposal,
  obs_interval = 7,
  time_start = 0,
  deterministic = FALSE
)
```

For the particle chains, acceptance is typically lower than the deterministic
pilot because each likelihood evaluation is noisy.

```{r synthetic-particle-acceptance}
acceptance_summary(synthetic_particle, "synthetic particle")
```

ESS and R-hat are complementary. ESS asks how many independent samples the
autocorrelated chain is worth; R-hat asks whether the different chains agree.

```{r synthetic-particle-ess-rhat}
ess_summary(synthetic_particle, "synthetic particle")
rhat_summary(synthetic_particle, "synthetic particle")
```

The synthetic parameters are recovered if the posterior intervals cover the
true values and the posterior mass is concentrated near them.

```{r synthetic-particle-summary}
parameter_summary(
  synthetic_particle,
  "synthetic particle",
  truth = truth_vec
)
```

```{r synthetic-particle-trace}
chlaa_plot_trace(
  synthetic_particle,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural"
)
```

```{r synthetic-particle-distributions}
chlaa_plot_parameter_distributions(
  synthetic_particle,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural",
  truth = truth_vec
)
```

```{r synthetic-particle-pairs, fig.height=7}
chlaa_plot_parameter_pairs(
  synthetic_particle,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural",
  truth = truth_vec,
  max_points = 3000
)
```

Finally, the posterior predictive check asks whether the fitted model can
regenerate data like the observed weekly counts.

```{r synthetic-fit-plot}
plot_case_fit(
  synthetic_particle,
  synthetic_weekly,
  "Posterior predictive fit to synthetic weekly cases",
  seed = 401
)
```

# Real Kirotshe fit

The real-data fit follows the same deterministic-to-particle sequence. There is
no known truth, so the emphasis shifts to chain agreement, predictive fit, and
whether the inferred parameter values are epidemiologically plausible.

```{r real-starts}
real_starts <- list(
  make_start(trans_prob = 2.0e-3, reporting_rate = 0.12, obs_size = 5, E0 = 600),
  make_start(trans_prob = 6.0e-4, reporting_rate = 0.60, obs_size = 50, E0 = 1000),
  make_start(trans_prob = 3.0e-3, reporting_rate = 0.20, obs_size = 10, E0 = 300)
)

imap_dfr(real_starts, \(pars, chain) tibble(
  chain = paste0("chain_", chain),
  parameter = natural_fit_names,
  value = unlist(pars[natural_fit_names])
)) |>
  pivot_wider(names_from = parameter, values_from = value)
```

## Deterministic fit

The real-data pilot again learns a proposal covariance with the deterministic
filter. This keeps the expensive particle-filter stage focused on a proposal
that already knows the posterior ridge.

```{r real-pilot, message=FALSE}
real_pilot <- chlaa_fit_pmcmc(
  data = fit_data_real,
  pars = real_starts[[1]],
  chain_pars = real_starts,
  n_chains = n_chains,
  n_particles = 1,
  n_steps = pilot_steps,
  seed = 501,
  prior = fit_prior,
  packer = make_packer(real_starts[[1]]),
  proposal_var = pilot_proposal,
  obs_interval = 7,
  time_start = 0,
  deterministic = TRUE
)
```

```{r real-pilot-acceptance}
acceptance_summary(real_pilot, "real pilot")
```

```{r real-det-proposal}
real_det_proposal <- proposal_from_fit(
  real_pilot,
  burnin = 0.25,
  scale = 1
)

round(cov2cor(real_det_proposal), 2)
```

```{r real-deterministic-fit, message=FALSE}
real_det <- chlaa_fit_pmcmc(
  data = fit_data_real,
  pars = real_starts[[1]],
  chain_pars = real_starts,
  n_chains = n_chains,
  n_particles = 1,
  n_steps = deterministic_steps,
  seed = 601,
  prior = fit_prior,
  packer = make_packer(real_starts[[1]]),
  proposal_var = real_det_proposal,
  obs_interval = 7,
  time_start = 0,
  deterministic = TRUE
)
```

```{r real-deterministic-diagnostics}
acceptance_summary(real_det, "real deterministic")
ess_summary(real_det, "real deterministic")
rhat_summary(real_det, "real deterministic")
```

```{r real-deterministic-summary}
parameter_summary(real_det, "real deterministic")
```

```{r real-deterministic-trace}
chlaa_plot_trace(
  real_det,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural"
)
```

```{r real-deterministic-pairs, fig.height=7}
chlaa_plot_parameter_pairs(
  real_det,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural",
  max_points = 3000
)
```

## Particle pMCMC

The real particle chains use 50 particles. The proposal is shrunk more than in
the synthetic example because the real likelihood is noisier and the posterior
is tighter around a low `obs_size`.

```{r real-particle-starts}
real_particle_starts <- chain_median_starts(
  real_det,
  template_pars = real_starts[[1]],
  burnin = 0.25
)
```

```{r real-particle-proposal}
real_particle_proposal <- proposal_from_fit(
  real_det,
  burnin = 0.25,
  scale = 0.2
)
```

```{r real-particle-fit, message=FALSE}
real_particle <- chlaa_fit_pmcmc(
  data = fit_data_real,
  pars = real_particle_starts[[1]],
  chain_pars = real_particle_starts,
  n_chains = n_chains,
  n_particles = particle_count,
  n_steps = particle_steps,
  seed = 701,
  prior = fit_prior,
  packer = make_packer(real_particle_starts[[1]]),
  proposal_var = real_particle_proposal,
  obs_interval = 7,
  time_start = 0,
  deterministic = FALSE
)
```

```{r real-particle-acceptance}
acceptance_summary(real_particle, "real particle")
```

```{r real-particle-ess-rhat}
ess_summary(real_particle, "real particle")
rhat_summary(real_particle, "real particle")
```

```{r real-particle-summary}
parameter_summary(real_particle, "real particle")
```

```{r real-particle-trace}
chlaa_plot_trace(
  real_particle,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural"
)
```

```{r real-particle-distributions}
chlaa_plot_parameter_distributions(
  real_particle,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural"
)
```

```{r real-particle-pairs, fig.height=7}
chlaa_plot_parameter_pairs(
  real_particle,
  parameters = natural_fit_names,
  burnin = 0.25,
  scale = "natural",
  max_points = 3000
)
```

The posterior predictive fit captures the broad two-wave structure. If the
highest late observations remain systematically above the median, that is a
model signal: the next model comparison should consider additional seeding,
initial-condition structure, or limited person-to-person transmission.

```{r real-fit-plot}
plot_case_fit(
  real_particle,
  kirotshe,
  "Posterior predictive fit to Kirotshe weekly cases",
  seed = 801
)
```

The fitted stochastic pMCMC object is saved as package example data. The
scenario and health-economics vignettes read this object, so they can start
from the fitted Kirotshe posterior rather than refitting the particle filter.

```{r save-fit-artifact}
fit_artifact <- list(
  fit = real_particle,
  pars = attr(real_particle, "start_pars", exact = TRUE),
  observed = kirotshe,
  interventions = kirotshe_meta,
  outbreak_start = outbreak_start,
  outbreak_end = outbreak_end,
  burnin = 0.25,
  particle_count = particle_count
)

artifact_path <- repo_output_file("inst", "extdata", "kirotshe_particle_fit.rds")
saveRDS(fit_artifact, artifact_path)
artifact_path
```

# Reading the diagnostics

The synthetic fit is the calibration check: the known truth should sit inside
the posterior intervals and near the high-density region of the marginal and
joint posterior plots. The real fit then asks a different question: whether the
small fitted model is stable enough to use as a baseline for scenario analysis.

The diagnostic tables and plots should be read together. Trace plots show
movement and chain overlap. ESS shows how much independent information the saved
draws contain. R-hat shows whether chains started from different points have
settled into the same distribution. Posterior predictive plots show whether
parameter uncertainty and observation noise generate outbreak curves like the
observed data.
