---
title: "Model Assumptions"
author: "acp25"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    embed-resources: true
---

## Purpose

This document audits the `chlaa` R package implementation against the reference paper and its supplementary materials:

> Loo PS, Rajah JK, Herrera De Leon HJ, Kopainsky B, Milano L (2025). "A simulation-based policy analysis of anticipatory action for cholera outbreaks, Democratic Republic of the Congo." *Bull World Health Organ* 103:607–618. doi:10.2471/BLT.25.293226

The supplementary materials (`.docx`) contain the complete Stella model documentation with all equations, stocks, flows, and parameter values. These are the primary source for structural comparison.

The goal is to confirm what is **faithfully reproduced**, what has been **deliberately adapted**, and what remains **uncertain or requiring further verification**.

------------------------------------------------------------------------

## 1. Model Architecture

### 1.1 Paper description

The paper uses a **system dynamics** (continuous, deterministic) SIRS-type model built in **Stella Architect v3.4.1**. It uses **Euler integration** with a time step of **dt = 0.25 days** over a 3-year period (Jan 2022 – Jan 2025). The population is not age- or sex-stratified.

The model has a rich compartmental structure including:

- **Latent compartment** (`recently_infected_population`) with `average_incubation_time = 3` days
- **Asymptomatic pathway**: `ratio_of_asymptomatic = 0.75`; the `asymptomatic_population` is an explicit stock with `average_duration_of_illness_asymptomatic = 7` days
- **Mild/severe severity split**: `normal_ratio_of_severe_disease = 0.30`; separate `mildly_infected_population` and `severe_infected_population` stocks
- **Treated/untreated distinction**: separate stocks for `treated_mildly_infected_population`, `treated_severe_infected_population`, `untreated_mildly_infected_population`, `untreated_severe_infected_population`
- **Separate recovered populations**: `recovered_asymptomatic_population`, `recovered_immune_treated_population`, `recovered_immune_untreated_population`
- **Immunity waning**: asymptomatic immunity 280 days (≈6 months in paper text); symptomatic immunity 1095 days (3 years)
- **Single vaccinated pool** with `vaccine_effectiveness = 0.76`, protection duration 180 days (1-dose) or 1095 days (2-dose)

Transmission is **exclusively indirect** (waterborne): *"Our model only incorporates indirect infection through contaminated water"* (Supplementary §1.1.1). The "contact rate" refers to contact with contaminated water, not person-to-person contact.

### 1.2 `chlaa` implementation

The package uses a **discrete-time stochastic** compartmental model compiled via **odin2/dust2**. Updates use **binomial draws** for integer-valued compartments, with a continuous environmental contamination state `C`. The time step is also **dt = 0.25**.


------------------------------------------------------------------------

## 2. Compartmental Structure

### 2.1 Comparison table

The paper's supplementary materials reveal that **most** of the compartmental structure in `chlaa` (latent, asymptomatic, mild/severe split) **already exists in the paper**. The key adaptations are in how compartments are grouped and how some paper features are simplified.

| Concept | Paper (Stella model) | `chlaa` (odin model) | Status |
|---|---|---|---|
| Susceptible | `susceptible_population` | `S` | **Faithful** |
| Latent/incubation | `recently_infected_population` (incubation 3 days) | `E` (incubation 4.845 days) | **Faithful** (parameter recalibrated) |
| Asymptomatic | `asymptomatic_population` (75%, duration 7 days) | `A` (75%, duration 5 days) | **Faithful** (parameter recalibrated) |
| Mild symptomatic (pre-triage) | `mildly_infected_population` | `M` | **Faithful** |
| Severe symptomatic (pre-triage) | `severe_infected_population` (30% of symptomatic) | `Sev` (30% of symptomatic) | **Faithful** |
| Mild untreated | `untreated_mildly_infected_population` | `Mu` | **Faithful** |
| Mild treated (ORC) | `treated_mildly_infected_population` | `Mt` | **Faithful** |
| Severe untreated | `untreated_severe_infected_population` | `Sevu` | **Faithful** |
| Severe treated (CTC) | `treated_severe_infected_population` | `Sevt` | **Faithful** |
| Recovered (asymptomatic) | `recovered_asymptomatic_population` (280 days) | `Ra` (odin default: 180 days) | **Faithful** (parameter discrepancy) |
| Recovered (treated sympt.) | `recovered_immune_treated_population` (1095 days) | Merged into `Rs` | **Simplified** |
| Recovered (untreated sympt.) | `recovered_immune_untreated_population` (1095 days) | Merged into `Rs` | **Simplified** |
| Vaccinated | Single `vaccinated_population` (effectiveness 0.76) | Dose-stratified `V1` (0.4), `V2` (0.7) | **Adapted** |
| Deaths (treated) | `treated_cholera_death` | `Dt` | **Faithful** |
| Deaths (untreated) | `untreated_cholera_death` | `Du` | **Faithful** |
| Environmental contamination | SMTHN(fraction_of_infected, time_to_contaminate, 3) | Explicit continuous state `C` with input/output ODE | **Adapted** (simplified dynamics) |
| CATI sub-model | 5 dedicated stocks + testing flows | Simple transmission multiplier (`cati_effect`) | **Heavily simplified** |
| Population with clean water | `population_access_water_system` (stock-based) | `chlor_effect` multiplier on transmission | **Heavily simplified** |
| Population with hygiene kits | `population_access_hygiene_kits` (stock-based) | `hyg_effect` multiplier on transmission | **Heavily simplified** |
| Latrine sufficiency | Stock-based construction + non-linear effect | `lat_effect` multiplier on shedding | **Heavily simplified** |
| ORC/CTC capacity | Dynamic goal-seeking with strain effects | Fixed capacity per day | **Simplified** |

### 2.2 Key structural differences

#### 2.2.1 Compartments already in the paper (NOT new to `chlaa`)

The following features were incorrectly attributed as `chlaa` innovations in earlier analysis. They are **present in the paper's Stella model**:

1. **Latent compartment**: The paper has `recently_infected_population` with `average_incubation_time = 3` days. `chlaa` uses `E` with recalibrated `incubation_time = 4.845` days.

2. **Asymptomatic pathway**: The paper explicitly models 75% asymptomatic infections (`ratio_of_asymptomatic = 0.75`) with a separate `asymptomatic_population` stock and `average_duration_of_illness_asymptomatic = 7` days. `chlaa` uses `prop_asym = 0.75` and `duration_asym = 5.0` days.

3. **Mild/severe split**: The paper has `normal_ratio_of_severe_disease = 0.30` with separate mild and severe stocks. `chlaa` uses `p_progress_severe = 0.30`.

4. **Separate recovered populations**: The paper distinguishes `recovered_asymptomatic_population`, `recovered_immune_treated_population`, and `recovered_immune_untreated_population`. `chlaa` merges the latter two into `Rs` (a minor simplification).

#### 2.2.2 Genuine structural adaptations in `chlaa`

1. **Stochastic discrete-time**: Paper uses deterministic continuous (Euler); `chlaa` uses binomial draws — a methodological upgrade.

2. **Dual-pathway transmission**: The paper uses **only indirect** (waterborne) transmission. `chlaa` adds an explicit **person-to-person** component:
   $$\lambda = \text{trans\_mult} \times \left( \underbrace{\text{contact\_rate} \times \text{trans\_prob} \times \frac{I_{\text{eff}}}{N}}_{\text{person-to-person (NEW)}} + \underbrace{\text{trans\_prob} \times \frac{C}{C + K}}_{\text{environmental}} \right)$$
   The paper's `contact_rate_with_contaminated_water = 6.204` refers to contact with water, not person-to-person contact.

3. **Environmental contamination dynamics**: The paper uses a third-order smooth delay (SMTHN) to model the fraction of contaminated water. `chlaa` replaces this with a simpler continuous ODE:
   $$\frac{dC}{dt} = \frac{\text{shed\_index}}{\text{time\_to\_contaminate}} - \frac{C}{\text{water\_clearance\_time}}$$
   The paper's approach normalises shedding against a `reference_CFU_cause_infection = 10^7` to produce a dimensionless "fraction of infected" which is then smoothed. `chlaa` instead uses a scaling factor (`contam_scale = 10^{10}`) and a half-saturation dose-response.

4. **Dose-stratified vaccination**: The paper uses a single `vaccinated_population` with `vaccine_effectiveness = 0.76` and switches between 1-dose (180 days) and 2-dose (1095 days) protection across scenario runs. `chlaa` implements `V1` and `V2` compartments with separate efficacies (`ve_1 = 0.4`, `ve_2 = 0.7`) allowing both regimens within a single model run.

5. **Simplified WASH interventions**: The paper models clean water and hygiene kit distribution as **stock-based interventions** — susceptible people flow into protected stocks (`population_access_water_system`, `population_access_hygiene_kits`) that reduce their contact rate by 70%. `chlaa` replaces this with simple multiplicative effect parameters (`chlor_effect`, `hyg_effect`) applied as `trans_mult = 1 - sum(active effects)`.

6. **Simplified CATI**: The paper has a complex CATI sub-model with 5 dedicated stocks (`CATI_recently_infected_population`, `CATI_asymptomatic`, `CATI_mildly_infected_population`, `CATI_severely_infected_population`, `CATI_targeted_population`) and testing flows that actively find and redirect infected individuals. `chlaa` reduces this to a single `cati_effect` transmission multiplier.

7. **Simplified ORC/CTC capacity**: The paper uses dynamic goal-seeking capacity building with construction delays (`time_to_build_ORC = 30` days) and **strain effects** — when facilities are overloaded, care quality drops (increasing progression from mild to severe) and fatality rates rise through non-linear functions. `chlaa` uses fixed capacity per day with no strain effects.

8. **Simplified latrine intervention**: The paper models latrines with stock-based construction, a `latrine_sufficiency` ratio, and a non-linear effect on contamination. `chlaa` uses a simple `lat_effect` shedding multiplier.

9. **Negative binomial observation model** for formal data fitting (entirely new).

10. **Health economics module** with costs and DALYs (entirely new).

------------------------------------------------------------------------

## 3. Parameter Audit

### 3.1 Supplementary base values vs `chlaa` defaults

The supplementary materials document explicit base values for all model parameters. The main paper's Table 1 lists 10 parameters varied ±15% in sensitivity analysis. Critically, **the midpoints of Table 1 ranges differ from the supplementary base values**, suggesting the model was recalibrated between versions.

```{r}
#| label: param-supplementary-vs-chlaa
#| echo: false
comparison <- data.frame(
    Parameter = c(
        "Contact rate",
        "Transmission probability",
        "Incubation time",
        "Duration asymptomatic illness",
        "Duration symptomatic illness",
        "Shedding: asymptomatic",
        "Shedding: mildly infected",
        "Shedding: severely infected",
        "Ratio of severe disease",
        "Ratio asymptomatic",
        "Asymptomatic immunity",
        "Symptomatic immunity",
        "Time to contaminate water",
        "Bacteria decay / clearance",
        "Fraction mild seeking care",
        "Fraction severe seeking care",
        "Treated fatality fraction",
        "Untreated fatality fraction",
        "Vaccine effectiveness",
        "1-dose protection",
        "2-dose protection"
    ),
    Supplementary = c(
        "6.204", "0.1278", "3 days", "7 days", "14 days",
        "100,000 CFU", "10,000,000 CFU", "30,000,000 CFU",
        "0.30", "0.75", "280 days", "1095 days", "14 days",
        "0.1 (10% decay)", "0.10", "0.20",
        "0.002", "0.005", "0.76", "180 days", "1095 days"
    ),
    Table1_Midpoint = c(
        "10.01", "0.055", "4.845 days", "-", "14.48 days",
        "90,690 CFU", "9,500,500 CFU", "32,945,000 CFU",
        "-", "-", "-", "-", "19.075 days",
        "-", "-", "0.68",
        "-", "-", "-", "-", "-"
    ),
    chlaa_odin = c(
        "10.01", "0.055", "4.845 days", "5.0 days", "14.48 days",
        "90,690 CFU", "9,500,500 CFU", "32,945,000 CFU",
        "0.30", "0.75", "180 days", "1095 days", "19.075 days",
        "30 days (clearance)", "0.30", "0.68",
        "0.0021", "0.50", "ve_1=0.4, ve_2=0.7", "180 days", "1095 days"
    ),
    chlaa_paramsR = c(
        "10.01", "0.127", "4.845 days", "5.0 days", "14.48 days",
        "90,690 CFU", "9,500,500 CFU", "32,945,000 CFU",
        "0.30", "0.75", "280 days", "1095 days", "19.075 days",
        "30 days (clearance)", "0.10", "0.20",
        "0.0021", "0.50", "ve_1=0.4, ve_2=0.7", "180 days", "1095 days"
    ),
    stringsAsFactors = FALSE
)

knitr::kable(comparison,
    caption = "Parameter comparison: Supplementary base values vs Main paper Table 1 midpoints vs chlaa defaults",
    col.names = c(
        "Parameter", "Supplementary base",
        "Table 1 midpoint", "chlaa odin default",
        "chlaa parameters.R"
    )
)
```

### 3.2 Key parameter findings

1. **Table 1 midpoints match chlaa odin defaults**: For all 10 sensitivity parameters in the main paper's Table 1, the midpoints of the ±15% ranges exactly match the `chlaa` odin model defaults. This suggests the model was recalibrated after the supplementary was written, and `chlaa` follows the final (published) calibration.

2. **Supplementary base values match `parameters.R` for some parameters**: `parameters.R` has `trans_prob = 0.127` (matching supplementary's 0.1278), `seek_mild = 0.10` (matching supplementary), `seek_severe = 0.20` (matching supplementary), and `immunity_asym = 280` (matching supplementary). This creates an internal inconsistency within `chlaa` — the odin model defaults follow the published Table 1, while `parameters.R` defaults follow the earlier supplementary values.

3. **Critical discrepancy — untreated fatality fraction**: The supplementary gives `untreated_fatality_fraction = 0.005` (0.5% CFR). `chlaa` uses `fatality_untreated = 0.50` (50% CFR) — a **100× difference**. The paper text states "untreated cholera can result in mortality rate as high as 70%", so the chlaa value (50%) is consistent with the paper's upper-bound statement but dramatically higher than the supplementary's calibrated base value. This is the most consequential parameter discrepancy and needs explicit justification.

4. **Environmental dynamics reparameterisation**: The paper uses `bacteria_decay = 0.1` (10% decay per day) and `time_to_contaminate_water = 14` days. `chlaa` uses `water_clearance_time = 30` days and `time_to_contaminate = 19.075` days. The environmental dynamics are structurally different (SMTHN vs simple ODE), so direct parameter comparison is not straightforward.

5. **Care-seeking probabilities diverge between chlaa's two sources**: The odin model uses `seek_mild = 0.30`, `seek_severe = 0.68` (matching Table 1); `parameters.R` uses `seek_mild = 0.10`, `seek_severe = 0.20` (matching supplementary). Which source is authoritative needs to be resolved.

### 3.3 Internal discrepancies: `parameters.R` vs `cholera_model.R`

```{r}
#| label: default-discrepancies
#| echo: false
disc <- data.frame(
    Parameter = c("trans_prob", "seek_mild", "seek_severe", "immunity_asym"),
    parameters.R = c(0.127, 0.10, 0.20, 280),
    cholera_model.R = c(0.055, 0.30, 0.68, 180),
    Suppl_Base = c(0.1278, 0.10, 0.20, 280),
    Table1_Midpoint = c(0.055, "N/A", 0.68, "N/A"),
    stringsAsFactors = FALSE
)

knitr::kable(disc,
    caption = "Internal defaults discrepancy: parameters.R follows supplementary, odin follows Table 1"
)
```

**Interpretation**: `parameters.R` appears to derive from an earlier version of the model (supplementary base values), while the odin model defaults follow the final published calibration (Table 1 midpoints). These need to be reconciled — users calling `chlaa_parameters()` will get different defaults from those baked into the odin model.

------------------------------------------------------------------------

## 4. Transmission Dynamics

### 4.1 Paper: indirect only

The supplementary explicitly states: *"Our model only incorporates indirect infection through contaminated water"* (§1.1.1).

The paper's infection rate is:

$$\text{infections} = S \times \underbrace{\text{fraction\_of\_infected\_contaminate\_water} \times \text{contact\_rate} \times \text{transmission\_probability}}_{\text{indicated\_infection\_rate}}$$

where `fraction_of_infected_contaminate_water` is a **third-order smooth delay** (`SMTHN`) of the normalised bacterial shedding. The normalisation divides total shedding by `reference_CFU_cause_infection = 10^7` to produce a dimensionless "fraction of infected".

### 4.2 `chlaa`: dual-pathway

`chlaa` adds a **person-to-person** transmission term alongside the environmental term:

$$\lambda = \text{trans\_mult} \times \left( \text{contact\_rate} \times \text{trans\_prob} \times \frac{I_{\text{eff}}}{N} + \text{trans\_prob} \times \frac{C}{C + K} \right)$$

**This is a deliberate extension** beyond the paper. In the paper, `contact_rate_with_contaminated_water = 6.204` describes contact with water, not with other people. The `chlaa` model reinterprets this as a general contact rate and adds person-to-person frequency-dependent transmission — standard in cholera literature (Codeco 2001, Tien & Earn 2010) but **not** in the reference paper.

### 4.3 Environmental contamination: SMTHN vs ODE

| Feature | Paper (Stella) | `chlaa` (odin) |
|---|---|---|
| Shedding normalisation | Total CFU / `reference_CFU = 10^7` → dimensionless | Total CFU / `contam_scale = 10^{10}` → index |
| Delay mechanism | SMTHN (3rd-order smooth, `time_to_contaminate = 14` days) | Simple ODE input/output (`time_to_contaminate = 19.075` days) |
| Decay | `bacteria_decay = 0.1` (10% per day) | `water_clearance_time = 30` days (first-order clearance) |
| Dose-response | Linear (`fraction × contact_rate × trans_prob`) | Saturating (`C / (C + K)` with `K = 1.0`) |
| Treated shedding | Treated at CTC/hospital: no environmental shedding | `treated_shed_mult_ctc = 0.0` (CTC), `treated_shed_mult_orc = 0.5` (ORC) |
| Latrine effect | Non-linear stock-based: `effect_of_latrine_sufficiency_on_contaminated_water` | Simple multiplier: `shed_mult = 1 - lat_active × lat_effect` |

------------------------------------------------------------------------

## 5. Intervention Structure

### 5.1 Interventions: paper vs `chlaa`

| Intervention | Paper (Stella) | `chlaa` (odin) | Faithfulness |
|---|---|---|---|
| **Clean water** | Stock-based: `population_access_water_system` with capacity building delays; reduces contact rate by 70% for protected individuals | `chlor_effect` multiplier on `trans_mult` (default 0.20) | **Heavily simplified** |
| **Hygiene kits** | Stock-based: `population_access_hygiene_kits` with capacity building delays; reduces contact rate by 70% for protected individuals | `hyg_effect` multiplier on `trans_mult` (default 0.20) | **Heavily simplified** |
| **Latrine construction** | Stock-based construction with `latrine_sufficiency` ratio and non-linear effect function on contamination (`baseline_effect = 1`, `latrine_sufficiency_effect = 0.5`) | `lat_effect` multiplier on `shed_mult` (default 0.10) | **Heavily simplified** |
| **CTC (severe cases)** | Dynamic capacity via goal-seeking; strain effects on fatality rate and care quality; initial capacity 20 persons | Fixed `ctc_capacity = 100` persons/day | **Simplified** (no strain) |
| **ORC (mild cases)** | Dynamic capacity via goal-seeking; strain effects on care quality (affects progression from mild→severe); initial capacity 50 persons | Fixed `orc_capacity = 500` persons/day | **Simplified** (no strain) |
| **CATI** | Complex sub-model with 5 stocks: `CATI_recently_infected`, `CATI_asymptomatic`, `CATI_mildly_infected`, `CATI_severely_infected`, `CATI_targeted`; testing capacity 19,900 over 60 days; diverts infected into intervention pathways | Single `cati_effect` multiplier on `trans_mult` (default 0.10) | **Heavily simplified** |
| **Vaccination** | Single pool; `vaccine_effectiveness = 0.76`; scenario switch for 1 vs 2 doses; 270,000 doses over 5-day campaign | Dose-stratified `V1`/`V2`; `ve_1 = 0.4`, `ve_2 = 0.7`; flexible campaign scheduling | **Adapted** (more flexible) |
| **Surveillance** | Modelled with `ratio_of_cases_being_reported = 0.7`; data collection with `normal_time_to_update_system = 14` days delay; affects intervention timing | Implicit via `reporting_rate = 0.2` in observation model | **Simplified** |

### 5.2 Paper's strain effects (absent from `chlaa`)

The paper models service strain through non-linear functions:

- **ORC strain** affects care-seeking rates and quality of care (mild → severe progression increases when ORCs are overloaded)
- **CTC strain** affects care-seeking and fatality rates (fatality fraction rises from `treated_fatality_fraction = 0.002` towards `service_strain_fatality_fraction = 0.0043` under strain)
- Both use sigmoidal functions with baseline effect 1.2, maximum strain effect 0.5, and capacity sensitivity exponent 2

`chlaa` omits all strain dynamics, using fixed capacity limits and constant fatality fractions. This is a significant simplification for outbreak scenarios where health system overload is expected.

### 5.3 Anticipatory action trigger

- **Paper**: Trigger at 15 cases/day, set by OCHA, based on real-time surveillance
- **`chlaa`**: `chlaa_trigger_time_from_sim()` finds first time `inc_symptoms` crosses a user-specified threshold
- **Status**: Faithful mechanism. The threshold is a user input, appropriate for generalisation.

------------------------------------------------------------------------

## 6. Scenario Structure

### 6.1 Paper scenarios vs `chlaa` defaults

| Paper Scenario | Description | `chlaa` equivalent | Status |
|---|---|---|---|
| Scenario 1: Baseline | Existing response (outbreak declaration Dec 2022, vaccination Jan 2023) | `"baseline"` | Faithful |
| Scenario 2: AA only | AA triggered Oct 2022, immediate WASH + CATI, same vaccination timeline | `"aa_no_vax"` | Faithful |
| Scenario 3: AA + 1 dose | AA + early vaccination (Nov 2022), 280k doses, 52% coverage, 6-month immunity | `"aa_vax_1dose"` | Faithful |
| Scenario 4: AA + 2 doses | AA + two-dose regimen, 280k doses to 140k people, 26% coverage, 3-year immunity | `"aa_vax_2dose"` | Faithful |
| — | — | `"no_intervention"` (counterfactual) | **New** |

### 6.2 Paper-specific values

- **Population**: N = 540,000 (Nyiragongo health zone) — `chlaa` default matches
- **Vaccine doses**: 270,000 (supplementary) / 280,000 (main paper text) — minor discrepancy
- **Vaccination campaign**: 5-day campaign, starting day 350 (Jan 2023 in baseline), 40 days after AA trigger in AA scenarios
- **Simulation period**: Jan 2022 – Jan 2025 (3 years = 1095 days)

------------------------------------------------------------------------

## 7. Health Economics

The paper **does not include health economic analysis**. The `chlaa` package adds a complete health economics module:

- **Costs**: vaccine delivery ($2/dose), ORC treatment ($10/case), CTC treatment ($80/case), WASH per person-day
- **DALYs**: disability weight 0.188 (GBD moderate diarrhoea), 7-day symptomatic duration, 30 YLL/death
- **Decision analysis**: CEAC, budget-constrained optimisation

**Status**: Entirely new. Cost parameters are labelled as "illustrative planning assumptions" in the references file.

------------------------------------------------------------------------

## 8. Observation Model

The paper validates against historical data using sensitivity analysis bands and a surveillance sub-model with `ratio_of_cases_being_reported = 0.7`. `chlaa` replaces this with a formal **negative binomial observation model**:

$$\text{cases} \sim \text{NegBin}(\mu = \text{reporting\_rate} \times \text{inc\_symptoms}, \text{size} = \text{obs\_size})$$

with defaults `reporting_rate = 0.2` and `obs_size = 25`.

**Status**: New capability. The paper's reporting rate (0.7) is very different from chlaa's (0.2) — this likely reflects different definitions (paper: fraction of symptomatic cases reaching health facilities and being recorded; chlaa: fraction of all symptomatic cases reported in surveillance data).

------------------------------------------------------------------------

## 9. Summary: Audit Classification

### Faithful to paper

1. Core SIRS compartmental framework with waning immunity
2. Latent compartment (`E` / `recently_infected_population`)
3. Asymptomatic pathway (75% of infections) with separate shorter immunity
4. Mild/severe severity split (30% severe) with separate treatment pathways
5. Treated/untreated distinction with separate death accounting
6. ORC (mild) and CTC (severe) care pathways
7. Key epidemiological parameters aligned to published Table 1 calibration
8. Vaccination with 180-day (1-dose) and 1095-day (2-dose) immunity
9. AA trigger mechanism (threshold-based on case counts)
10. 4-scenario structure (baseline, AA, AA+1dose, AA+2dose)
11. Time step dt = 0.25
12. Population size N = 540,000

### Deliberately adapted

1. **Stochastic discrete-time** (binomial draws) instead of deterministic continuous (Euler)
2. **Person-to-person transmission added** — the paper is purely waterborne
3. **Environmental contamination simplified** from SMTHN (3rd-order smooth) to simple input/output ODE with half-saturation dose-response
4. **Dose-stratified vaccination** (`V1`, `V2` with separate efficacies) instead of single pool with effectiveness 0.76
5. **Merged recovered populations** — paper separates treated/untreated recovered; chlaa merges into `Rs`
6. **WASH interventions simplified** from stock-based (people flow into protected stocks, 70% contact rate reduction) to simple transmission multipliers (20% reduction each)
7. **CATI simplified** from complex 5-stock sub-model with testing flows to single transmission multiplier
8. **ORC/CTC capacity simplified** from dynamic goal-seeking with strain effects to fixed capacity
9. **Latrine intervention simplified** from stock-based construction with non-linear sufficiency effect to simple shedding multiplier
10. **ORC partial shedding** (`treated_shed_mult_orc = 0.5`) — paper treats all ORC-treated as reducing contamination but through care-seeking rather than explicit shedding reduction
11. **Negative binomial observation model** for formal fitting (new)
12. **Health economics and optimisation module** (new)
13. **"No intervention" counterfactual scenario** (new)

### Uncertain / requiring verification

1. **`parameters.R` vs odin model defaults**: `trans_prob` (0.127 vs 0.055), `seek_mild` (0.10 vs 0.30), `seek_severe` (0.20 vs 0.68), `immunity_asym` (280 vs 180) — appears to be supplementary vs published calibration; needs reconciliation
2. **Untreated fatality fraction**: Supplementary = 0.005 (0.5%), chlaa = 0.50 (50%) — 100× difference. Paper text says "as high as 70%"; supplementary base value is much lower. Which is correct for the target context?
3. **Person-to-person transmission**: Not in the paper at all. What is the intended relative contribution vs environmental? Should `contact_rate` be reinterpreted?
4. **WASH effect sizes** (20% each for chlor/hygiene, 10% latrine/CATI): Paper uses 70% contact rate reduction for clean water/hygiene kits and a non-linear latrine effect — very different magnitudes
5. **Additive vs multiplicative intervention effects**: `trans_mult = 1 - sum(effects)` is additive; paper uses stock-based individual-level protection
6. **Half-saturation dose-response** (`contam_half_sat = 1.0`) and **contam_scale** (`1e10`): Calibration targets? The paper uses a normalised linear dose-response, not saturating
7. **Water clearance time** (30 days): Paper uses `bacteria_decay = 0.1` (10%/day ≈ 10-day half-life). The 30-day clearance in `chlaa` is substantially slower
8. **Reporting rate discrepancy**: Paper surveillance model = 0.7; chlaa = 0.2
9. **Strain effects omission**: The paper's non-linear strain on fatality and care quality is absent — how much does this affect outbreak dynamics when health systems are overloaded?
10. **Model calibration provenance**: Were the Table 1 parameter values (which differ from supplementary) obtained by fitting to Nyiragongo data? If so, do they implicitly account for structural simplifications?

------------------------------------------------------------------------

## 10. Recommended Actions

### Immediate (before next fitting run)

1. **Reconcile parameter defaults**: Align `parameters.R` with `cholera_model.R`. Decide whether to follow supplementary base values or published Table 1 calibration. Document the decision.
2. **Investigate untreated fatality fraction**: The 100× discrepancy (0.005 vs 0.50) is consequential. The supplementary value of 0.005 (0.5% CFR) reflects North-Kivu context; chlaa's 0.50 (50%) represents severe untreated cholera at its worst. These have very different implications for intervention impact estimates.
3. **Document person-to-person pathway**: Clarify in `model_mapping.md` that this is an extension beyond the paper, and assess its contribution to total force of infection.

### Before publication

4. **Review WASH effect magnitudes**: The paper's 70% contact rate reduction for clean water/hygiene is much larger than chlaa's 20% default. Consider whether the simplified multiplier approach can capture similar effect sizes.
5. **Assess strain effects**: Run sensitivity analysis to determine whether omitting ORC/CTC strain dynamics materially changes conclusions about intervention impact.
6. **Validate against Nyiragongo data**: Replicate the paper's Fig. 3 historical validation.
7. **Document economic assumptions**: Current unit costs are illustrative; replace with DRC-specific values.

------------------------------------------------------------------------

## 11. Updated Modelling Assumptions Register

```{r}
#| label: assumptions-register
#| echo: false
register <- data.frame(
    ID = paste0("A", sprintf("%02d", 1:22)),
    Assumption = c(
        "Homogeneous, well-mixed population (no age/sex/spatial structure)",
        "Stochastic discrete-time dynamics with binomial transitions",
        "75% of infections are asymptomatic (faithful to paper)",
        "30% of symptomatic infections progress to severe (faithful to paper)",
        "Asymptomatic immunity lasts 180 days (odin) or 280 days (parameters.R)",
        "Symptomatic immunity lasts 3 years (1095 days) — faithful to paper",
        "Dual-pathway transmission: person-to-person + environmental",
        "Environmental contamination follows simple input/output ODE (paper uses SMTHN)",
        "Half-saturation dose-response for environmental transmission (paper is linear)",
        "Care-seeking: 30% mild, 68% severe (odin) or 10% mild, 20% severe (parameters.R)",
        "CTC eliminates environmental shedding (treated_shed_mult_ctc = 0)",
        "ORC reduces environmental shedding by 50% (treated_shed_mult_orc = 0.5)",
        "Case fatality: 0.21% treated, 50% untreated (paper supplementary: 0.2% treated, 0.5% untreated)",
        "Fixed ORC/CTC capacity with no strain effects (paper has dynamic capacity and strain)",
        "WASH interventions modelled as simple multipliers (paper uses stock-based protection)",
        "WASH default effects: chlor 20%, hygiene 20%, latrine 10%, CATI 10% (paper: 70% contact reduction)",
        "CATI modelled as transmission multiplier (paper has 5-stock sub-model with testing)",
        "Vaccine efficacy: 40% (dose 1), 70% (dose 2) — paper: single pool at 76%",
        "Negative binomial observation model with 20% reporting rate (paper: 70%)",
        "AA trigger is threshold-based on symptomatic incidence",
        "No seasonality or environmental forcing",
        "No population displacement, conflict, or demographic change"
    ),
    Source = c(
        "Paper (explicit simplification)",
        "Adaptation from paper (paper uses deterministic SD)",
        "Paper supplementary: ratio_of_asymptomatic = 0.75",
        "Paper supplementary: normal_ratio_of_severe_disease = 0.30",
        "Supplementary: 280d; odin default: 180d; paper text: '3 to 12 months'",
        "Paper: 'immunity typically lasts 3 years'",
        "Extension: paper is waterborne only (supplementary §1.1.1)",
        "Adaptation: paper uses SMTHN third-order smooth delay",
        "Assumption: paper uses linear fraction × contact_rate × trans_prob",
        "Odin matches Table 1; parameters.R matches supplementary base",
        "Paper: treated waste disinfected at hospital/CTC",
        "Assumption: ORC = community-level, less controlled waste management",
        "Paper suppl: untreated_fatality_fraction = 0.005; chlaa = 0.50 (100× higher)",
        "Simplification: paper has goal-seeking capacity with strain functions",
        "Simplification: paper models individual-level stock-based protection",
        "Paper: 70% contact rate reduction for clean water/hygiene kit recipients",
        "Simplification: paper's CATI has testing, identification, referral pathways",
        "Adaptation: enables both regimens in single model run",
        "New for fitting; paper surveillance reports 70% of cases",
        "Paper: 15 cases/day threshold set by OCHA",
        "Paper: 'did not consider seasonal factors'",
        "Paper: 'did not consider population displacement, conflict'"
    ),
    Confidence = c(
        "High", "High (methodological choice)", "High (matches paper)",
        "High (matches paper)", "Low (internal discrepancy)",
        "High", "Medium (standard in cholera lit, but extends paper)",
        "Medium (different dynamics, needs calibration comparison)",
        "Low (paper uses linear; calibration implications unclear)",
        "Low (internal discrepancy; 0.10/0.20 or 0.30/0.68?)",
        "High", "Low (needs source)",
        "Low (100× discrepancy with supplementary needs resolution)",
        "Medium (may underestimate outbreak severity under overload)",
        "Medium (captures direction but not magnitude of paper's effects)",
        "Low (magnitudes much smaller than paper's 70%)",
        "Medium (captures direction; loses CATI's testing/identification mechanism)",
        "High (improves on paper's scenario-switch approach)",
        "Low (very different from paper's 70%)",
        "High",
        "High (explicit limitation)", "High (explicit limitation)"
    ),
    stringsAsFactors = FALSE
)

knitr::kable(register,
    caption = "Modelling assumptions register (updated Meeting 4)"
)
```

------------------------------------------------------------------------

## 12. Conclusion


`chlaa` makes **significant structural simplifications** in three domains:

1. **Interventions**: The paper's complex stock-based WASH interventions (70% contact rate reduction), dynamic capacity building with strain effects, and CATI sub-model are all replaced by simple multipliers or fixed capacities. This is the largest area of divergence.

2. **Transmission**: `chlaa` adds person-to-person transmission (absent from the paper) and simplifies environmental contamination dynamics from a third-order smooth delay to a simple ODE with half-saturation dose-response.

